### 2016-07-01 finalText/proposedSol.tex
Preliminary Thesis Version Completed

\subsection{Finite-precision representation system}
Before discussing any details concerned with the actual hardware implementation, I will layout some of the design choices that were made regarding to how we numbers envolved in the calculations will be represented. For that purpose, Section~\ref{sec:precbit} discusses the fundamentals of fixed-point representation systems, as well as the bitwidth and precision chosen for then number representation system of the network, and Section~\ref{sec:convrulesfp} states how the conversion between real and fixed-point numbers is performed. Finally, Section~\ref{sec:arithrulesfp} explains the special rules that fixed-point arithmetic imposes. The considerations made in this chapter can be found on~\cite{Yates13}, which is a good reference for fixed-point arithmetic theory.

\subsubsection{Precision bitwidth}\label{sec:precbit} 
Since we are dealing with real numbers, and I plan to make use of the DSP48 slices within the FPGA, I chose to use an 18-bit signed fixed-point system, with the sign information coded as 2's complement. Fixed-point systems are usually addressed in the $Qn.m$ form, where $n$ is the bitwidth of the of the integer part (excluding the sign bit) and $m$ is the bitwidth of the fractional part, and so the total bitwidth is $N=m+n+1$ to account for the sign bit. This way, the value of the $i$-th position bit is $2^{i-m}$, and since this is a 2's complement system, the last bit is $-2^{n-1}$. In terms of range of representation, the maximum positive number that can be represented corresponds to all bits set to one, except for the last ($2^{n+m+1}-1$), but shifted by $m$ bits to the right to yield the correct real number (the decimal point is at the $m$-th bit),

@@

Assuming that the smallest perturbation used in the training system will be $2^{-9}$, according to the previous Python experiments, our fractional part precision should be, at least, greater than this, so a sensible choice would be either $m=10$ or $m=11$. On the other hand, given that in the Python experiments I have attested that the weight values generaly do not (and should not) exceed values around the first power of ten, there is no need for large values of $n$ (although it is advisable to not be very small in order to accomodate the intermediate calculations, avoiding overflow), and so we should choose to have as much precision as possible, so $m=11$. Since $18=n+m+1$, then $n = 17 - m = 6$ and the representation system to be used is $Q6.11$. According to Equations~\ref{eq:maxpos} and~\ref{eq:minneg}, the real values $x$ than can be represented with this choice of $n$ and $m$ are in the range

@@

\subsubsection{Theoretical considerations on the approximation method}
The polynomial approximation methods aim to approximate some function $f(x)$ in an interval $\left[a, b\right]$ using a polynomial $p^*_n \in \mathcal{P}$ of degree $n$, in order to meet the optimization criteria chosen \textit{a priori}. This optimization criteria can be either the well-known \textbf{Least Squares Approximation} procedure, where we minimize the \emph{average quadratic error} $\left[f(x)-p^*(x)\right]^2$, or the \textbf{Least Maximum Approximation}, where we \emph{minimize} the \emph{maximum possible error}, also commonly called a \textit{minimax} approximation. Since we are operating in $Q6.11$ fixed-point arithmetic, we need to guarantee that the maximum approximation error does not exceed the precision limit of this representation, $2^{-11}$, and so a \textit{minimax} approach is desirable, since it guarantees that a given maximum error is not exceeded. Weierstrass's Theorem~\cite{Weierstrass1885}, from 1885, guarantees that there is always a polynomial that can approximate any continuous function f with error $\epsilon > 0$. Chebyshev also proved~\cite{Muller05} that, in a \textit{minimax} polynimal approximation of degree $n$, the minimum approximation error $\epsilon$ is achieved in at least $n+2$ points, and the sign of approximation error alternates from one interval to the other, and thus the error is not \emph{biased}. This leads to a linear system of $n+2$ equations whose $i$-th line is given by

@@

The optimal coefficients of this \textit{minimax} polynomial are found using \textit{Remez Algorithm}, which provides an iterative approach to solve the linear system given by Equation~\ref{eq:remezline} by finding, in each iteration, the $n+2$ set of points $x_i$ of Chebyshev's Theorem that minimize the error function to $\epsilon$. The algorithm operations are as follows

\begin{enumerate}
    \item Initializing the set $x_i$ of points to $x_i = \frac{a+b}{2} + \frac{(b-a)}{2}\cos\left(\frac{i\pi}{n+1}\right), \, 0 \leq i \leq n+1$ -- 
    \item Solve the system in~\ref{eq:remezline}
    \item Given the polynomial coefficients yielded by step 2, compute the $y_i$ points that minimize $p(x)-f(x)$, and replace the $x_i$s of the next iteration by these $y_i$s. Go to step 2 until $\epsilon$ does not decrease.
\end{enumerate}

@@

\begin{equation}
\begin{bmatrix} 1 & x_0 & x_0^2 & \cdots & x_0^n & - 1 \\ 1 & x_1 & x_1^2 & \cdots & x_1^n & + 1 \\  &  &  & \vdots & & \\ 1 & x_{n+1} & x_{n+1}^2 & \cdots & x_{n+1}^n & (-1)^{n+1} \end{bmatrix}
\begin{bmatrix} p_0 \\ p_1 \\ p_2 \\ \vdots \\ p_n \\ \epsilon \end{bmatrix} = \begin{bmatrix} f(x_0) \\ f(x_1) \\ \vdots \\ f(x_{n+1}) \end{bmatrix}
\end{equation}
and therefore the solution vector $p = \left[p_0 \, p_1 \, p_2 \, \cdots \, p_n \, \epsilon \right]$ for Step 2 of Remez's Algorithm is simply given by $p = A^{-1}b$, where $A$ is the $x_i$s matrix and $b$ is the $f(x_i)$s vector. The Python implementation of this algorithm is presented in Listing~\ref{lst:remezpy}, as follows
\lstinputlisting[language=Python, caption=Python script that implements Remez's algorithm, label={lst:remezpy}]{remez.py}

Instead of using a \emph{single} polynomial of higher degree ($n \geq 3$) for the whole domain, I chose to partition the domain of the activation functions in \textbf{6 intervals}, and approximate each of those intervals using polynomials of degree $n=2$. This proved to yield a lower overall approximation error, as expected (the interval on which to perform the approximation is smaller). Also, since both the $\sigma(x)$ and $\tanh(x)$ have horizontal assymptotes in $\{0,1\}$ and $\{-1,1\}$ respectively, the far-left and far-right intervals do not need a polynomial approximation, and can be assigned a constant value equal to the corresponding assymptote. The resulting \textit{minimax} approximation polynomials yielded by running the code in Listing~\ref{lst:remezpy} are

@@

\subsubsection{Hardware Implementation}
The hardware module that implements these non-linearities is, essentially, a 2nd degree polynomial calculator, where the coefficients of the polynomial are chosen accordingly with the value of the input $x$. This last functionality
is implemented using a simple multiplexer that loads the signals $p_0$, $p_1$ and $p_2$ with the coefficients of Equations~\ref{eq:coefs_sigm} and~\ref{eq:coefs_tanh} based on the value of the input operand $x$. As for the polynomial calculator is concerned, although we could use a full-pipelined evaluator, that would require two DSP slices -- to perform the two simultaneous multiplications -- but the DSP slices will be heavily used in the matrix-vector calculators, so it is advisable to save them for that purpose. A simpler approach is to note that by factoring $x$ we get

\begin{equation}\label{eq:factorPol}
p(x) = p_0 + p_1x + p_2x^2 = p_0 + x(p_1 + xp_2)
\end{equation}
where we can note that this operation can be divided in a two-step procedure of a simultaneous mutiplication of the operand by a constant, and a subsequent addition of another constant: first, by multiplying the operand by $p_2$ and summing $p_1$, and then by multiplying the operand by this last result and summing $p_0$. The block diagram of the hardware implementation of the non-linearity calculator is presented in Figure~\ref{fig:nonlin}. Also, in Figure~\ref{fig:nonlin-out}, the output of the Verilog module that implements this design is compared with the actual output (I used Numpy as reference) for both activation functions.

@@

\subsection{Matrix-vector Dot Product Unit}\label{sec:dotprod_sec}
From the set of Equations~\ref{eq:equationsLSTM}, we see that the weight matrices $\mb{W}_*$ and $\mb{R}_*$ are multiplied by the input vector $\mb{x}$ and the layer output vector $\mb{y}$, respectively. This way, we need a HDL block that implements matrix-vector multiplication, and that block must be parameterized in order to accomodate different matrices/vectors of various sizes: note that $x$ has length $M$ (the number of inputs to the layer), and so $\mb{W}_*$ has size $N\times M$, while $y$ has lenght $N$ (the number of neurons in the layer), and so $\mb{R}_*$ now has size $N\times N$. Plus, if we need a layer with different parameters, either in terms of number of inputs or number of neurons, it advisable to only change the respective parameter at the synthesis stage instead of having to redesign the whole block for that particular size.

@@

Although this solution only requires one multiplication per row of the input matrix (i.e. $N$ multiplications), if the row size is large, we may run out of resources in the FPGA; therefore, some sort of \textit{resource multiplexing} strategy must be used to ensure the flexibility of the solution to accomodate networks of larger dimensions. The solution I have found for this issue was to \emph{share} the multiplication slice between rows of the matrix: before, each multiplication slice was responsible for producing the $i$-th element of the output vector $y$ (of size $N$), therefore the final result for the vector would be ready in $M$ clock cycles (i.e. the number of columns); now, if we define a parameter $K_G = \frac{\text{Number of rows}}{\text{Number of multipliers}}$ -- the number of rows that share the same multiplier -- the same multiplier is responsible for producing several $i$-th elements of the output vector, in consecutive time slots of $M$ clock cycles. Suppose that we have a $8\times2$ matrix, and that $K_G = 2$; this way, we would have 4 multipliers, and the output vector elements $y_0$, $y_2$, $y_4$ and $y_6$ would be ready after $M=2$ clock cycles, and the remaining -- $y_1$, $y_3$, $y_5$ and $y_7$ --  are ready after another two clock cycles, that is $2M = 4$ clock cycles after the calculations began.

In Figures-\ref{fig:mem-arrayprod} and~\ref{fig:array-prod} are depicted a diagram of the memory access for the Matrix, and the row multiplication and units within the module, respectively, where I have set $K_G = 4$, for the same matrix and vector sizes as before. Note that in this situation, we would only have 2 multipliers, and the module would be composed of two multiplication units, such as those in Figure~\ref{fig:array-prod}, that work in parallel: they address a particular column using the signal \verb+colSel+, which is used by the RAM module to output the corresponding column of the matrix (in regard to the input vector, obviously this signal selects only a single position), depicted in Figure~\ref{fig:mem-arrayprod}. The dark shaded part of the memory is used by the first multiplier, and the light shaded is used by the other, in parallel for a fixed \verb+rowMux+ -- this signal is produced by the control unit of the module, and essentially operates the left Mux and right Demux of Figure~\ref{fig:array-prod} that allows to choose the proper position of the weight column and to write to the correct output register, respectively. In this example, for \verb+rowMux+=0, the control unit increments \verb+colSel+ from 0 to $M$, and thus evaluating $y_0$ and $y_4$. After this, \verb+rowMux+ is incremented to 1, and the process repeats until \verb+rowMux+ reaches $K_G-1$.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/array-prod.eps}
    \caption[The $i$-th row multiplication unit of the Module]{The $i$-th row multiplication unit of the Module, where rowMux and colSel are internal signals produced by the control unit of the module. The flip-flop accumulates the sum, and the output demux selects the appropriate memory position on where to store this value, within the slot attributed to this multiplication, from $i\cdot K_G$ to $i\cdot K_G + \left[K_G-1\right]$}
    \label{fig:array-prod}
\end{figure}
In summary, for an $N \times M$ weight matrix, this module outputs the result vector in $K_G \cdot M$ clock cycles.

\subsection{Dual-port Block RAM}
Although the bias weights, being simply a vector, I have chosen to keep them in normal registers and let the synthesis tool decide how to place them in the FPGA, for the weight matrices, the case is different. Since the
network size can be big, it is good to make sure that they are placed in the Block RAMs available in the FPGA. In Section~\ref{sec:dotprod_sec}, we see that, because of the architecture of the matrix-vector multiplier, it is convenient to have a column by column access to the weight matrix, and therefore this block outputs a given column in the \verb+rowOut+ output port terminal, at the \emph{negative edge} of the clock, selected by the value present in the input \verb+addressOut+ at the immediately previous \emph{positive edge} of the clock. As far as \textbf{writing} to the memory is concerned, the process is identical but now the column of weights at the input \verb+rowIn+ is sampled at \emph{negative edge} of the clock, and is placed at the column specified by the input \verb+addressIn+ at the previous \emph{positive edge} of the clock.

@@

@@

\begin{enumerate}
    \item Multiply matrix $\mb{W}_*$ by the input vector $\mb{x}^{(t)}$
    \item Multiply matrix $\mb{R}_*$ by the previous layer output vector $\mb{y}^{(t-1)}$
    \item Sum the bias vector $\mb{b}_*$ to the remaining matrix-vector dot product results.
\end{enumerate}
Assuming that the network size $N$ is always larger than the input size $M$, if we use the matrix-vector dot product units of Section~\ref{sec:dotprod_sec}, the multiplication in task 1 takes $K_G\cdot M$ cycles and the one in task 2 takes $K_G\cdot N$ cycles. This way, task 2 and task 1 can be performed in parallel, and we can use the extra time that task 2 takes, relative to task 1,  to perform task 3, and sum the bias vector to the output of task 1, whose result is ready by that time.

@@

\section{Hardware LSTM Network -- Forward Propagation}\label{sec:hLSTM-fp}
After detailing the design of each of the constituent modules of the LSTM Network, I will now show how they are interconnected
to implement a hardware version of Equations~\ref{eq:equationsLSTM}. This will be performed in Section~\ref{sec:hLSTM-struct}, where the design decisions and compromises that were taken will be detailed, along with some estimates on the resource usage and calculation time for each of the design iteration versions, the naïve solution (~\ref{sec:struct-fullpar}) and an optimized version of the former (~\ref{sec:struct-shared}) that exploits some of the hardware redundancies by multiplexing in time the usage of those redundant structures.

@@

\begin{equation}\label{eq:numdsp_network}
    4\frac{2N}{K_G} + 5N + 3N = N \left( \frac{8}{K_G} + 8 \right).
\end{equation}
In terms of time performance, we can measure it by estimating the number of clock cycles needed to perform a complete forward propagation. Since the gate module outputs a result in $K_G \cdot N$ cycles, the activation function evaluators need always 2, and the elementwise multiplicators need only 1 cycles, and noting that the two elementwise multiplicators that sum to produce $\mb{c}^{(t)}$ can work in parallel, the estimated number of clock cycles needed is

\begin{equation}\label{eq:numcc_network}
    (N \cdot K_G + 2) + (2 + 1) + (2 + 1) = 8 + N\cdot K_G
\end{equation}

@@

\subsubsection{Shared Elementwise Multiplication Block}\label{sec:struct-fullpar}
In Figure~\ref{fig:network}, we see that, apart from the elementwise multiplicator preceeding the flip-flops, all of them follow a similar structure: one of the operands is the output of a $\tanh(\mb{x})$ block and the other from a $\sigma(\mb{x})$. A clever improvement over the last network architecture is to instead of replicating these $\tanh$-$\sigma$-$(\cdot)$wise structures use a \emph{single one} and choose the operand accordingly to the state that the network is currently in. Besides, the right elementwise multiplier of Figure~\ref{fig:network} is not used as the same time as any other, so it is a perfect waste of resources. The issue about the elementwise multiplier that precedes the flip-flops can be solved by adding another multiplexer that chooses between the output of the $\tanh(\mb{x})$ module or the signal $\mb{c}^{(t-1)}$. These ideas resulted in the improved network design of Figure~\ref{fig:network-opt}.

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{figures/network-opt.eps}
    \caption[Optimised version of~\ref{fig:network} that shares the elementwise multiplication and the activation function blocks]{Optimised version of~\ref{fig:network} that shares the elementwise multiplication and the activation function blocks}
    \label{fig:network-opt}
\end{figure}
The two left multiplexers control the operands that are fed to the activation function modules, and the selecting signal is generated by the network's state machine, and its value is incremented after each complete usage of the $\tanh$-$\sigma$-$(\cdot)$wise  structure: this is where the time multiplexing of the structure takes place. Since in state $\verb+Sel+=1$ the left operand of the elementwise multiplicator (the one that preceeded the flip-flops in the previous design) is the signal $\mb{c}^{(t-1)}$, I added another multiplexer before the elementwise multiplication that selects that signal in that particular case, and the output from the $\tanh(\mb{x})$ block, otherwise.

The flip-flops on the righthand side of Figure~\ref{fig:network-opt} are activated by signals generated within the network state machine that enable the appropriate flip-flop, placing the output from the elementwise multiplicator in the correct place. The first activated flip-flop is the middle one, which keeps the result from the operation of the $\mb{z}^{(t)}$ and $\mb{i}^{(t)}$ vectors, then, after a full operation of the elementwise multiplier, the bottom flip-flop to save the other portion of the sum that evaluates to the $\mb{c}^{(t)}$ signal. Lastly, the top flip-flop saves the network output $\mb{y}^{(t)}$, which in the next incoming sample becomes $\mb{y}^{(t-1)}$ and is used by the Gate modules in this next batch of calculations.

>>>>

\subsection{Finite-precision representation system}
Before discussing any other details concerned with the actual hardware implementation, I will layout some of the design choices that were made regarding how the numbers involved in the calculations are represented. For that purpose, Section~\ref{sec:precbit} discusses the fundamentals of fixed-point representation systems, as well as the bitwidth and precision chosen for then number representation system of the network, and Section~\ref{sec:convrulesfp} states how the conversion between real and fixed-point numbers is performed. Finally, Section~\ref{sec:arithrulesfp} explains the special rules that fixed-point arithmetic imposes. The considerations made in this chapter can be found on~\cite{Yates13}, which is a good reference for fixed-point arithmetic theory.

\subsubsection{Precision bitwidth}\label{sec:precbit} 
Since we are dealing with real numbers, and I plan to make use of the DSP48 slices within the FPGA, I chose to use an 18-bit signed fixed-point system, with the sign information coded as 2's complement. Fixed-point systems are usually addressed in the $Qn.m$ form, where $n$ is the bitwidth of the of the integer part (excluding the sign bit) and $m$ is the bitwidth of the fractional part, and so the total bitwidth is $N=m+n+1$ to account for the sign bit. In this way, the value of the $i$-th position bit is $2^{i-m}$, and since this is a 2's complement system, the last bit is $-2^{n-1}$. In terms of range of representation, the maximum positive number that can be represented corresponds to all bits set to one, except for the last ($2^{n+m+1}-1$), but shifted by $m$ bits to the right to yield the correct real number (the decimal point is at the $m$-th bit),

@@

Assuming that the smallest perturbation used in the training system will be $2^{-9}$, according to the previous Python experiments, our fractional part precision should be, at least, greater than this, so a sensible choice would be either $m=10$ or $m=11$. On the other hand, given that in the Python experiments I have attested that the weight values generally do not (and should not) exceed values around the first power of ten, there is no need for large values of $n$ (although it is advisable to be large enough to accommodate the intermediate calculations, avoiding overflow), and so we should choose to have as much precision as possible, so $m=11$. Since $18=n+m+1$, then $n = 17 - m = 6$ and the representation system to be used is $Q6.11$. According to Equations~\ref{eq:maxpos} and~\ref{eq:minneg}, the real values $x$ than can be represented with this choice of $n$ and $m$ are in the range

@@

\subsubsection{Theoretical considerations on the approximation method}
The polynomial approximation methods aim to approximate some function $f(x)$ in an interval $\left[a, b\right]$ using a polynomial $p^*_n \in \mathcal{P}$ of degree $n$, in order to meet the optimization criteria chosen \textit{a priori}. This optimization criteria can be either the well-known \textbf{Least Squares Approximation} procedure, where we minimize the \emph{average quadratic error} $\left[f(x)-p^*(x)\right]^2$, or the \textbf{Least Maximum Approximation}, where we \emph{minimize} the \emph{maximum possible error}, also commonly called a \textit{minimax} approximation. Since we are operating in $Q6.11$ fixed-point arithmetic, we need to guarantee that the maximum approximation error does not exceed the precision limit of this representation, $2^{-11}$, and so a \textit{minimax} approach is desirable, since it guarantees that a given maximum error is not exceeded. Weierstrass's Theorem~\cite{Weierstrass1885}, from 1885, guarantees that there is always a polynomial that can approximate any continuous function f with error $\epsilon > 0$. Chebyshev also proved~\cite{Muller05} that, in a \textit{minimax} polynomial approximation of degree $n$, the minimum approximation error $\epsilon$ is achieved in at least $n+2$ points, and the sign of approximation error alternates from one interval to the other, and thus the error is not \emph{biased}. This leads to a linear system of $n+2$ equations whose $i$-th line is given by

@@

The optimal coefficients of this \textit{minimax} polynomial are found using the \textit{Remez Algorithm}, which provides an iterative approach to solve the linear system given by Equation~\ref{eq:remezline} by finding, in each iteration, the $n+2$ set of points $x_i$ of Chebyshev's Theorem that minimize the error function to $\epsilon$. The algorithm operations are as follows

\begin{enumerate}
    \item Initializing the set $x_i$ of points to $x_i = \frac{a+b}{2} + \frac{(b-a)}{2}\cos\left(\frac{i\pi}{n+1}\right), \, 0 \leq i \leq n+1$ 
    \item Solve the system in~\ref{eq:remezline}
    \item Given the polynomial coefficients yielded by step 2, compute the $y_i$ points that minimize $p(x)-f(x)$, and replace the $x_i$s of the next iteration by these $y_i$s. Go to step 2 until $\epsilon$ does not decrease.
\end{enumerate}

@@

\begin{equation}
\begin{bmatrix} 1 & x_0 & x_0^2 & \cdots & x_0^n & - 1 \\ 1 & x_1 & x_1^2 & \cdots & x_1^n & + 1 \\  &  &  & \vdots & & \\ 1 & x_{n+1} & x_{n+1}^2 & \cdots & x_{n+1}^n & (-1)^{n+1} \end{bmatrix}
\begin{bmatrix} p_0 \\ p_1 \\ p_2 \\ \vdots \\ p_n \\ \epsilon \end{bmatrix} = \begin{bmatrix} f(x_0) \\ f(x_1) \\ \vdots \\ f(x_{n+1}) \end{bmatrix}
\end{equation}
and therefore the solution vector $p = \left[p_0 \, p_1 \, p_2 \, \cdots \, p_n \, \epsilon \right]$ for Step 2 of Remez's Algorithm is simply given by $p = A^{-1}b$, where $A$ is the matrix on the left-hand side of the equation and $b$ is the vector on the right-hand side. The Python implementation of this algorithm is presented in Listing~\ref{lst:remezpy} of Appendix~\ref{ap1:remez-pyth}.

Instead of using a \emph{single} polynomial of higher degree ($n \geq 3$) for the whole domain, I chose to partition the domain of the activation functions in \textbf{6 intervals}, and approximate each of those intervals using polynomials of degree $n=2$. This proved to yield a lower overall approximation error, as expected (the interval on which to perform the approximation is smaller). Also, since both the $\sigma(x)$ and $\tanh(x)$ have horizontal asymptotes in $\{0,1\}$ and $\{-1,1\}$ respectively, the far-left and far-right intervals do not need a polynomial approximation, and can be assigned a constant value equal to the corresponding asymptote. The resulting \textit{minimax} approximation polynomials yielded by running the code in Listing~\ref{lst:remezpy} are

@@

\subsubsection{Hardware Implementation}
The hardware module that implements these non-linearities is, essentially, a 2nd degree polynomial calculator, where the coefficients of the polynomial are chosen accordingly with the value of the input $x$. This last functionality
is implemented using a simple multiplexer that loads the signals $p_0$, $p_1$ and $p_2$ with the coefficients of Equations~\ref{eq:coefs_sigm} and~\ref{eq:coefs_tanh} based on the value of the input operand $x$. As for the polynomial calculator is concerned, although we could use a full-pipelined evaluator, that would require two DSP slices -- to perform the two simultaneous multiplications -- but the DSP slices will be heavily used in the matrix-vector calculators, so it is advisable to save them for that purpose. A simpler approach is to note that, according to Horner's Rule, we get

\begin{equation}\label{eq:factorPol}
p(x) = p_0 + p_1x + p_2x^2 = p_0 + x(p_1 + xp_2)
\end{equation}
where we can note that this operation can be divided in a two-step procedure of a simultaneous multiplication of the operand by a constant, and a subsequent addition of another constant: first, by multiplying the operand by $p_2$ and summing $p_1$, and then by multiplying the operand by this last result and summing $p_0$. The block diagram of the hardware implementation of the non-linearity calculator is presented in Figure~\ref{fig:nonlin}. Also, in Figure~\ref{fig:nonlin-out}, the output of the Verilog module that implements this design is compared with the actual output (I used Python3's Numpy as reference) for both activation functions.

@@

\subsection{Matrix-vector Dot Product Unit}\label{sec:dotprod_sec}
From the set of Equations~\ref{eq:equationsLSTM}, we see that the weight matrices $\mb{W}_*$ and $\mb{R}_*$ are multiplied by the input vector $\mb{x}$ and the layer output vector $\mb{y}$, respectively. This way, we need a HDL block that implements matrix-vector multiplication, and that block must be parameterized in order to accommodate different matrices/vectors of various sizes: note that $x$ has length $M$ (the number of inputs to the layer), and so $\mb{W}_*$ has size $N\times M$, while $y$ has length $N$ (the number of neurons in the layer), and so $\mb{R}_*$ now has size $N\times N$. Plus, if we need a layer with different parameters, either in terms of number of inputs or number of neurons, it advisable to only change the respective parameter at the synthesis stage instead of having to redesign the whole block for that particular size.

@@

Although this solution only requires one multiplication per row of the input matrix (i.e. $N$ multiplications), if the row size is large, we may run out of resources in the FPGA; therefore, some sort of \textit{resource multiplexing} strategy must be used to ensure the flexibility of the solution to accomodate networks of larger dimensions. The solution I have found for this issue was to \emph{share} the multiplication slice between rows of the matrix: in a direct implementation of Algorithm~\ref{matvec-alg}, each multiplication slice was responsible for producing the $i$-th element of the output vector $y$ (of size $N$), therefore the final result for the vector would be ready in $M$ clock cycles (i.e. the number of columns); now, if we define a parameter $K_G = \frac{\text{Number of rows}}{\text{Number of multipliers}}$ -- the number of rows that share the same multiplier -- the same multiplier is responsible for producing several $i$-th elements of the output vector, in consecutive time slots of $M$ clock cycles. Suppose that we have a $8\times2$ matrix, and that $K_G = 2$; in this way, we would have 4 multipliers, and the output vector elements $y_0$, $y_2$, $y_4$ and $y_6$ would be ready after $M=2$ clock cycles, and the remaining -- $y_1$, $y_3$, $y_5$ and $y_7$ --  are ready after another two clock cycles, that is $2M = 4$ clock cycles after the calculations began.

Figures-\ref{fig:mem-arrayprod} and~\ref{fig:array-prod} depict a diagram of the memory access for the Matrix, and the row multiplication and units within the module, respectively, where I have set $K_G = 4$, for the same matrix and vector sizes as before. Note that in this situation, we would only have 2 multipliers, and the module would be composed of two multiplication units, such as those in Figure~\ref{fig:array-prod}, that work in parallel: they address a particular column using the signal \verb+colSel+, which is used by the RAM module to output the corresponding column of the matrix (in regard to the input vector, obviously this signal selects only a single position), depicted in Figure~\ref{fig:mem-arrayprod}. The dark shaded part of the memory is used by the first multiplier, and the light shaded is used by the other, in parallel for a fixed \verb+rowMux+ -- this signal is produced by the control unit of the module, and essentially operates the left Mux and right Demux of Figure~\ref{fig:array-prod} that allows to choose the proper position of the weight column and to write to the correct output register, respectively. In this example, for \verb+rowMux+=0, the control unit increments \verb+colSel+ from 0 to $M$, and thus evaluating $y_0$ and $y_4$. After this, \verb+rowMux+ is incremented to 1, and the process repeats until \verb+rowMux+ reaches $K_G-1$. Therefore, we have the correct result vector in a time proportional to $K_G \cdot M$: since the memory only outputs the appropriate column in the \emph{next clock cycle}, and that this module is pipelined both at the input and at the output in order to increase the maximum clock frequency, we should add a 3 clock cycle overhead to the previous estimate.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/array-prod.eps}
    \caption[The $i$-th row multiplication unit of the Module]{The $i$-th row multiplication unit of the Module, where rowMux and colSel are internal signals produced by the control unit of the module. The flip-flop accumulates the sum, and the output demux selects the appropriate memory position on where to store this value, within the slot attributed to this multiplication, from $i\cdot K_G$ to $i\cdot K_G + \left[K_G-1\right]$}
    \label{fig:array-prod}
\end{figure}

\subsection{Dual-port weight RAM}
Even though I have chosen to keep the bias weight vectors in normal registers and let the synthesis tool decide how to place them in the FPGA, for the weight matrices, the case is different. Since the network size can be large, it is good to make sure that they are placed in the RAMs available in the FPGA. In Section~\ref{sec:dotprod_sec}, we see that, because of the architecture of the matrix-vector multiplier, it is convenient to have a column by column access to the weight matrix, and therefore this block outputs a given column in the \verb+rowOut+ output port terminal, at the \emph{negative edge} of the clock, selected by the value present in the input \verb+addressOut+ at the immediately previous \emph{positive edge} of the clock. As far as \textbf{writing} to the memory is concerned, the process is identical but now the column of weights at the input \verb+rowIn+ is sampled at \emph{negative edge} of the clock, and is placed at the column specified by the input \verb+addressIn+ at the previous \emph{positive edge} of the clock.

@@

After performing synthesis, I have noticed that using block RAM led to a slightly poorer performance, so I have settled with the use of \emph{Distributed RAM}, storing the weight matrices in LUTRAM.

@@

\begin{enumerate}
    \item Multiply matrix $\mb{W}_*$ by the input vector $\mb{x}^{(t)}$
    \item Multiply matrix $\mb{R}_*$ by the previous layer output vector $\mb{y}^{(t-1)}$
    \item Sum the bias vector $\mb{b}_*$ to the remaining matrix-vector dot product results.
\end{enumerate}
Assuming that the network size $N$ is always larger than the input size $M$, if we use the matrix-vector dot product units of Section~\ref{sec:dotprod_sec}, the multiplication in task 1 takes approximately $K_G\cdot M$ cycles and the one in task 2 takes $K_G\cdot N$ cycles. This way, task 2 and task 1 can be performed in parallel, and we can use the extra time that task 2 takes, relative to task 1,  to perform task 3, and sum the bias vector to the output of task 1, whose result is ready by that time. The module is triggered by a \verb+beginCalc+ input signal that activates the internal state machine, and outputs a \verb+dataReady+ signal that informs the network that the calculations have been concluded, and the value at the output is the actual final result. 
After validating and simulating this block, and taking into account the internal state-machine and that the internal datapath is pipelined, the exact number of clock cycles this module takes to output a result is $6+K_G\cdot N$.

@@

\section{Hardware LSTM Network -- Forward Propagation}\label{sec:hLSTM-fp}
After detailing the design of each of the constituent modules of the LSTM Network, I will now show how they are interconnected
to implement a hardware version of Equations~\ref{eq:equationsLSTM}. This is performed in Section~\ref{sec:hLSTM-struct}, where the design decisions and compromises that were taken are detailed, along with some estimates on the resource usage and calculation time for each of the design iteration versions, the naïve solution (~\ref{sec:struct-fullpar}) and an optimized version of the former (~\ref{sec:struct-shared}) that exploits some of the hardware redundancies by multiplexing in time the usage of those redundant structures.

@@

\begin{equation}\label{eq:numdsp_network}
    4\frac{2N}{K_G} + 5N + 3N = N \left( \frac{8}{K_G} + 8 \right).
\end{equation}
In terms of time performance, we can measure it by estimating the number of clock cycles needed to perform a complete forward propagation. Since the gate module outputs a result in $K_G \cdot N$ cycles, the activation function evaluators need always 5, and the elementwise multiplicators need only 2 cycles (one cycle for the pipeline, and another for the actual calculation), and noting that the two elementwise multiplicators that sum to produce $\mb{c}^{(t)}$ can work in parallel, the estimated number of clock cycles needed would be proportional to

\begin{equation}\label{eq:numcc_network}
    (N \cdot K_G + 6) + (5 + 2) + (5 + 2) = 20 + N\cdot K_G
\end{equation}

@@

\subsubsection{Shared Elementwise Multiplication Block}\label{sec:struct-shared}
In Figure~\ref{fig:network}, we see that, apart from the elementwise multiplier preceeding the flip-flops, all of them follow a similar structure: one of the operands is the output of a $\tanh(\mb{x})$ block and the other from a $\sigma(\mb{x})$. An improvement over the last network architecture is to instead of replicating these $\tanh$-$\sigma$-$(\cdot)$wise structures use a \emph{single one} and choose the operand accordingly to the state that the network is currently in. Besides, the right elementwise multiplier of Figure~\ref{fig:network} is not used as the same time as any other, so it is a perfect waste of resources. The issue about the elementwise multiplier that precedes the flip-flops can be solved by adding another multiplexer that chooses between the output of the $\tanh(\mb{x})$ module or the signal $\mb{c}^{(t-1)}$. These ideas resulted in the improved network design of Figure~\ref{fig:network-opt}.

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{figures/network-opt.eps}
    \caption[Optimised version of~\ref{fig:network} that shares the elementwise multiplication and the activation function blocks]{Optimised version of~\ref{fig:network} that shares the elementwise multiplication and the activation function blocks}
    \label{fig:network-opt}
\end{figure}
The two left multiplexers control the operands that are fed to the activation function modules, and the selecting signal is generated by the network's state machine, and its value is incremented after each complete usage of the $\tanh$-$\sigma$-$(\cdot)$wise  structure: this is where the time multiplexing of the structure takes place. Since in state $\verb+Sel+=1$ the left operand of the elementwise multiplicator (the one that preceded the flip-flops in the previous design) is the signal $\mb{c}^{(t-1)}$, I added another multiplexer before the elementwise multiplication that selects that signal in that particular case, and the output from the $\tanh(\mb{x})$ block, otherwise.

The flip-flops on the right hand side of Figure~\ref{fig:network-opt} are activated by signals generated within the network state machine that enable the appropriate flip-flop, placing the output from the elementwise multiplicator in the correct place. The first activated flip-flop is the middle one, which keeps the result from the operation of the $\mb{z}^{(t)}$ and $\mb{i}^{(t)}$ vectors, then, after a full operation of the elementwise multiplier, the bottom flip-flop to save the other portion of the sum that evaluates to the $\mb{c}^{(t)}$ signal. Lastly, the top flip-flop saves the network output $\mb{y}^{(t)}$, which in the next incoming sample becomes $\mb{y}^{(t-1)}$ and is used by the Gate modules in this next batch of calculations.

Now, since there is only a single elementwise multiplier and only two activation function calculators, the total requirement for DSP slices is simply
\begin{equation}\label{eq:numdsp_network-opt}
    4\frac{2N}{K_G} + 2N + N = N \left( \frac{8}{K_G} + 3 \right).
\end{equation}
where we see that we saved $5N$ multipliers, which for a large value of $N$ can have a decisive impact. In terms of speed performance, although the Gate calculation time remains the same, now the $\tanh$-$\sigma$-$(\cdot)$wise structure runs for 3 consecutive times, in a non parallel fashion. After adjustments to the state machine, and accounting for pipelining and synchronization within the datapath, the clock cycles needed after the gate calculations are 27, so the total clock cycles needed are

\begin{equation}\label{eq:numcc_network-opt}
    (N \cdot K_G + 6) + 27  = 33 + N\cdot K_G
\end{equation}
which is only 13 clock cycles more than the fully-parallel architecture. For instance, an $N=32$ neuron network would require 320 DSP slices, while this new architecture would only require 160, only at the expense of 13 more clock cycles.
